Variational sampling reloaded

### 2024-05-03 papers/composite_inference/notes2024.tex
More notes on VS

\def\x{{\mathbf{x}}}
\def\y{{\mathbf{y}}}
\newcommand{\blambda}{{\boldsymbol{\lambda}}}
\newcommand{\Blambda}{{\boldsymbol{\Lambda}}}
\newcommand{\bell}{{\boldsymbol{\ell}}}
\newcommand{\E}{\mathbb{E}}

@@

The idea of variable feature (VF) based likelihood was introduced by Baggenstoss in \cite{Baggenstoss-03}. It is definitely a relevant idea, and perhaps an important one too, but this initial work was not fully successful.

On the positive side, Baggenstoss gave a simple and general way, which he called ``PDF~projection theorem'', to construct a VF-based likelihood function. As later noted by \cite{Minka-04}, it is just an application of the maximum entropy, or minimum KL~divergence, principle.

However, it is easy to see how the PDF~projection method can lead to heavily biased, hence completely absurd, likelihood functions, if the variable features are chosen arbitrarily. Fixed features can be chosen arbitrarily -- at worst, they will lead to a flat likelihood, but can't bias inference. The same cannot be said for variable features.

Baggenstoss probably saw that, but brushed the issue under the carpet: ``The choice of features remains an art requiring intuition''. This is nonsense. The choice of features should be driven by explicit assumptions.

The PDF~projection method simply needs an additional constraint to work. I have known the solution for quite some time but have struggled to fully contemplate it.

\subsection{The missing constraint}

@@

>>>>

\def\x{{\mathbf{x}}}
\def\y{{\mathbf{y}}}
\def\z{{\mathbf{z}}}
\newcommand{\blambda}{{\boldsymbol{\lambda}}}
\newcommand{\Blambda}{{\boldsymbol{\Lambda}}}
\newcommand{\bell}{{\boldsymbol{\ell}}}
\newcommand{\E}{\mathbb{E}}

@@

The idea of variable feature (VF) likelihood was introduced by Baggenstoss in \cite{Baggenstoss-03} under the not very explicit name of ``class-specific method''. It is definitely a relevant idea, but this initial work was not fully successful for several reasons.

On the positive side, Baggenstoss gave a simple and general way, which he called ``PDF~projection theorem'', to construct a VF-based likelihood function. As later noted by \cite{Minka-04}, it is just an application of the maximum entropy or, more generally, the minimum KL~divergence principle.

However, there are three main flaws in Baggenstoss's presentation in my humble (or not so much) opinion:
\begin{enumerate}
\item {\em Bias}. It is easy to see how the PDF~projection method can lead to heavily biased likelihood functions if the variable features are chosen arbitrarily. The truth is that fixed features can be chosen arbitrarily -- at worst, they will lead to a flat likelihood, but can't bias inference. The same cannot be said for variable features. Baggenstoss may have seen that, but brushed the issue under the carpet: ``The choice of features remains an art requiring intuition''. Still, the assumptions under which a particular choice of features works or not should be made explicit.
\item {\em Arbitrary reference hypothesis}. The PDF~projection theorem involves a reference, or null hypothesis distribution that Baggenstoss gives no guideline whatsoever on how to choose. Since the likelihood function that arises from PDF~projection clearly depends on the reference distribution, this choice matters. He tried to work around this issue by making the reference distribution parametric, which led him to the third flaw:
\item {\em Inconsistent parameter elimination}. Baggenstoss treated the VF likelihood exactly as if it was a true likelihood function, and implicitely proposed to eliminate the nuisance parameters by maximization. The deep information-theoretic interpretation of PDF~projection shows that it is wrong. It is easy to see that Baggenstoss's approach doesn't work in the simplest instance of VF likelihood where the parameter space is a singleton -- the method should then be equivalent to the well-known maximum likelihood ratio method for binary hypothesis testing. This suggests that the nuisance parameters pertaining to the null hypothesis cannot be just mixed with those of the feature distribution; instead, they should be considered as competitive parameter sets.
\end{enumerate}

\subsection{Solving for the bias: the missing constraint}

The PDF~projection method simply needs an additional assumption to be unbiased. I have known the solution for quite some time but have struggled to fully contemplate it.

@@

\appendix

\section{Baggenstoss 2003 paper: reading notes}

He introduces the class-specific approach by a curse of dimensionality argument:
\begin{quote}
Because each feature set can be designed without regard to other classes (or states), it can be of much lower dimension than a common feature set that must account for all classes, effectively avoiding the curse of dimensionality.
\end{quote}

This is spot on: we want to select features to avoid the curse of dimensionality, however class-independent feature selection may not give us enough flexibility. He further comments on that:
\begin{quote}
[Feature selection] leads to a fundamental tradeoff: whether to discard features in an attempt to reduce the dimension to something manageable or to include them and suffer the problems associated with estimating a PDF at high dimension.
\end{quote}

He mentions that he did publish a previous work on the topic in 1999, which was based on sufficient statistics and used a fixed reference hypothesis.

His exposition of the PDF~projection theorem starts with the remark that, if $X$ is a random vector and $Z = f(X)$ is a one-to-one function of~$X$, then:
$$
p_x(\x)
=
|J_f(\x)| p_z(\z)
.
$$

In the more general case of a many-to-one function, there are multiple distributions~$p_x(\x)$ that are compatible with a given~$p_z(\z)$. To solve for this ambiguity, Baggenstoss postulates that a reference distribution is needed:
\begin{quote}
In order to apply the constraint, it is necessary to make use of a reference hypothesis $H_0$ for which we know the PDF of both $\x$ and $\z$.
\end{quote}

In his derivation of the PDF~projection theorem, he postulates the

### 2024-05-03 papers/variana/2024.tex
More notes on VS

\title{Variana Reloaded}
\author{Alexis Roche}
%\date{}

@@

@@

It is time to assemble the pieces of the puzzle.
\begin{itemize}
\item 2010: discovery of the VS core idea and its intriguing exactness property.
\item 2011--13: consolidation of the theory behind VS.
\item 2014--16: realization that VS is useful in practice combined with Expectation Propagation.
\item 2018--19: extension to large dimension via custom optimizer (avoiding big Hessian inversion).
\item 2022: proven bound on VS asymptotic variance.
\item 2023: final key insight: VS makes optimization possible in high dimension (at least locally) because it avoids saddle points unlike gradient-based methods.
\end{itemize}

@@

@@

Next, let us consider the loss function $\ell(f,g)=f\log(f/g)+g-f$ as a function of $g$. We easily see that:
$$
\frac{\partial\ell}{\partial g} = 1 - \frac{f}{g},
\qquad
\frac{\partial^2\ell}{\partial g^2} = \frac{f}{g^2},
$$
implying that the loss is not only convex in~$g$ but also staisfies the conditions of Alexis's theorem with $a=f$. Hence,
$$
\forall g\leq f,
\qquad
\ell(f, g) \geq \frac{1}{2f} (g - f)^2. 
$$

@@

Integrating this over~$\pi$, we get this long-waited inequality:
$$
\int \pi (f-g)^2 \leq 2\max(\|f\|_\infty, \|g\|_\infty) D(\pi f\|\pi g),
$$
where the infinite norms are taken on the support of~$\pi$, which is infinite if $\pi$ is a Gaussian kernel.

@@

>>>>

\title{Variational sampling reloaded}
\author{Alexis Roche}
%\date{}

@@

\section{Introduction}

Variational sampling is a simple and general method for computational statistics that extends in a rather natural way the basic importance sampling method, aka Monte Carlo integration, which dates back (at least) to World War II.

I discovered it in 2010 while trying to interpret importance sampling in a variational framework, and coined the name ``variational sampling'' as a result of that. Despite its relative simplicity, it did not seem to be known under any name in the literature, which puzzled me back then, and keeps puzzling me.

There could only be two reasons for that. Either it was a bad idea, or researchers had overlooked it. I was crazy enough to believe in the latter explanation.

Hence, I tried hard to publish it in NIPS, but failed four times in as many attempts from 2010 to 2016. I did publish it on arXiv in May 2011 on the advice of Pr Nanny Wermuth, who had come across my work and encouraged me to push it forward, and later in a low cost e-journal nicely called {\em International Journal of Advanced Statistics and Probability} in a desperate, yet successful, attempt to make at least one peer-reviewed publication on the topic.

As of May 2024, these papers have a total of 4 citations in Google Scholar, including 2 self-citations.

Variational sampling is, unquestionably, the worst failure of my whole researcher career in terms of scientific impact, that being said while said career operates on reasonably modest standards.

%%%, but much simpler than many advanced sampling and approximate inference methods found in the literature.
%thought it should be part of any standard computational statistics toolbox

@@

How the pieces of the puzzle came together.
\begin{itemize}
\item 2010: discovery of the VS core idea and its intriguing exactness property.
\item 2011--13: consolidation of the theory behind VS.
\item 2014--16: realization that VS is useful in practice combined with Expectation Propagation.
\item 2018--19: extension to large dimension via custom optimizer (avoiding big Hessian inversion); unveiling connection between distribution scaling and $\alpha$-divergence minimization.
\item 2022: proven bound on VS asymptotic variance.
\item 2023: revisited custom optimizer.
\end{itemize}

@@

Note, however, that IS can also be seen as a variational sampling instance corresponding to another surrogate criterion:
$$
\tilde{D}_{IS}(\theta)
= - \frac{1}{n} \sum_i f(x_i) \log g_\theta(x_i) 
  + \int \pi g_\theta
.
$$

The key difference with $\tilde{D}(\theta)$ is that the sample mean of $g_\theta(x_i)$ is replaced by its asymptotical value. Hence, $\tilde{D}_{IS}$ may be seen as a ``partial Rao-Blackwellization'' of $\tilde{D}$, partial because we can't do it for the term in $\log g_\theta(x_i)$; and, because it's partial, it may not improve the estimator accuracy (our main claim is that it doesn't in many situations of practical interest).

@@

Next, let us consider the loss function $\ell(f,g)=f\log(f/g)+g-f$ as a function of $g$. We easily see that:
$$
\frac{\partial\ell}{\partial g} = 1 - \frac{f}{g},
\qquad
\frac{\partial^2\ell}{\partial g^2} = \frac{f}{g^2},
$$
implying that the loss is not only convex in~$g$ but also satisfies the conditions of Alexis's theorem with $a=f$. Hence,
$$
\forall g\leq f,
\qquad
\ell(f, g) \geq \frac{1}{2f} (g - f)^2. 
$$

@@

Integrating this over~$\pi$, we get this long-waited inequality:
\begin{equation}
\label{eq:variance_inequality}
\int \pi (f-g)^2 \leq 2\max(\|f\|_\infty, \|g\|_\infty) D(\pi f\|\pi g),
\end{equation}
where the infinite norms are taken on the support of~$\pi$, which is infinite if $\pi$ is a Gaussian kernel.

@@

Substituing~$g$ with the minimizer~$g_\star$ of $D(\pi f\|\pi g)$ in the inequality~(\ref{eq:variance_inequality}), we get an upper bound for the VS 0-order moment estimate asymptotic variance:
$$
\int \pi (f-g_\star)^2
\leq
2\max(\|f\|_\infty, \|g_\star\|_\infty)
D(\pi f\|\pi g_\star)
.
$$

Obviously, a sufficient condition for the left-hand side to be finite is that the right-hand side be finite.

We note that, if $\|f\|_\infty$ is finite, then $D(\pi f\|\pi g_\star)$ is finite. The reason is that, by construction, $g_\star$ minimizes the KL~divergence on a function space that contains the constant function $g\equiv 1$, therefore:
$$
D(\pi f\|\pi g_\star)
\leq
D(\pi f\|\pi 1)
=
\int \pi (f\log f - f + 1)
.
$$

Now, the function $x\log x - x + 1$ is positive-valued on $\mathbb{R}_+$, decreasing from 1 to 0 between 0 and 1, and increasing to infinity from 1. It follows that $(f\log f - f + 1)\leq \max(1, \|f\|_\infty \log \|f\|_\infty - \|f\|_\infty + 1)$, hence the KL~divergence term is finite since it is upper bounded by a finite constant:
$$
D(\pi f\|\pi g_\star) \leq
\max(1, \|f\|_\infty \log \|f\|_\infty - \|f\|_\infty + 1)
.
$$

Is the condition $\|f\|_\infty$ finite sufficient?

For the IS estimate, it is -- because the asymptotic variance is related to the variance of $f(x)$ under $\pi(x)$:
$$
{\rm Var}[f(X)]
= \min_{c\in\mathbb{R}} \int \pi (f - c)^2,
$$
which is easily seen to be finite whenever $f$ is upper bounded.

For VS, however, we need $\|g_\star\|_\infty$ to be finite too. Intuitively, since $g_\star$ is the best fit to $f$ in the KL~sense, the residuals $f-g_\star$ tend to be much smaller than wrt a constant baseline, so we expect $\int \pi(f-g_\star)^2$ to be quite smaller than ${\rm Var}[f(X)]$. We can make this a little bit more precise using :
$$
\min_g \int \pi (f-g)^2 \leq \min_{c\in\mathbb{R}} \int \pi (f-c)^2 = {\rm Var}[f(X)].
$$

However, $g_\star$ does not minimize $\pi (f-g)^2$, it minimizes the KL divergence $D(\pi f\|\pi g)$, therefore the above reasoning doesn't work.

So let's go back to the KL~minimization. One thing we can note is that the variances of $\pi f$ and $\pi g_\star$ are the same owing to the moment matching property of KL~minimization, and so are the precisions:
$$
\prec(\pi g_\star) = \prec(\pi f).
$$

Now, both $\pi$ and $g_\star$ are log-quadratic functions so the precisions add up:
$$
\prec(\pi g_\star) = \prec(\pi) + \prec(g_\star).
$$

Clearly, $\|g_\star\|_\infty$ is finite iff $\prec(g_\star)$ is positive (not necessarily positive definite). This is the case iff: 
$$
\prec(\pi f) \succeq \prec(\pi)
.
$$

In other words, sufficient conditions for VS to converge are:
\begin{enumerate}
\item The factor $f$ is upper bounded.
\item The ``posterior'' variance is smaller than the ``prior'' variance.
\end{enumerate}

Is the second condition implied by the first? No, it isn't.

\paragraph{Damning counter-example.}

Here is a simple one-dimensional counter-example where $\|f\|_\infty$ is finite and $\|g_\star\|_\infty$ is not. Consider a normalized Gaussian reference $\pi = N(0,1)$, and a uniform factor defined by $f = \chi(-a,a)/\pi$, where $a$ is a positive constant, so that the target distribution $\pi f$ is the uniform distribution $\chi(-a,a)$ and its variance is therefore $a^2/ 3$. Following the same reasoning as above, we see that the precision of the fitted factor is:
$$
\prec(g_\star) = \frac{3}{a^2} - 1,
$$
which, for $a$ sufficiently large, namely $a>\sqrt{3}$, is strictly negative, implying that $g_\star$ is not upper bounded. Even worse than that, $g_\star$ goes very fast to infinity as $x$ departs from $0$, so $\int \pi g_\star^2$ is infinite, implying that $\int \pi (f-g_\star)^2$ is also infinite, implying that VS does not converge!

In this case, the optimal fit $g_\star$ goes to infinity while $f$ vanishes. This is weird. But it happens because, on the one hand, the factor looks like a U-shaped function in the context defined by $\pi$, and, on the other hands, the loss incurred in regions where the factor vanihes, $\ell(0,g)=g$, is only linear in~$g$, so it's ``mild'' even for large values of~$g$ when weighted by~$\pi$. This is consistent with the fact that the KL~divergence is an inclusive  measure (``zero avoiding'' for $g_\star$). In contrast, minimizing $\pi(f-g)^2$ would yield a well-behaved (bell-shaped) factor approximation -- to see this, just realize that $\inf_g\int\pi(f-g)^2 \leq \int\pi f^2$, which is finite.

\paragraph{Wrapping up.}

We learn several things from the above counter-example.

First off, VS is not necessarily better than IS (it may not converge at all while IS converges).

Second, the intuition that $g_\star$ follows $f$ closely is not necessarily correct.

Third, this intuition may not be correct even though the factor is ``almost'' perfectly log-quadratic on the domain defined by $\pi$. In the counter-example, VS will do a good job in practice (much better than IS). Why? Just think about it: because on a finite sample drawn from $\pi$, chances are that we don't get any point outside $[-a,a]$, where $f$ is perfectly log-quadratic.

%For instance, drawing 100000 points from a normal distribution, we have almost no chance to have any point more than 5 times standard deviations away from the mean. With 1000000 points, the same goes for 6 standard deviations.

Hence, and that's a 4th take-away, the asymptotic variance does not tell us the whole truth about how VS performs in practice. The sample size required to reach asymptotic regime may heavily depend on the factor at hand. Besides, the fact that VS may actually worsen by increasing the sample size is both unexpected and concerning.

As already pointed out, truncating the reference distribution doesn't really solve the problem. The asymptotic variance would become finite, but still possibly way too big. In the counter-example, how would we know beforehand that we should truncate outside $[-a,a]$?

So, how can we work around this problem?

%we can put the blame on both the factor for being far off the approximating family and the context $\pi$ as for being too narrow to ``see'' $f$ globally. But we may not know that beforehand.

%Then comes a natural question: when is the variance of $(f-g_\star)$ smaller than that of $f$ under~$\pi$?

\paragraph{Bullshite.}
